---
title: "who’s tuberculosis analysis"
output: pdf_document
---

"who" dataset contains tuberculosis (TB) cases broken down by year, country, age, gender, and diagnosis method.
This dataset is very typical real-life example. It contains redundant columns, odd variable codes and many missing values. We need to tidy this messy dataset.

we have:
- `country`, `iso2`, `iso3` are variables
- `year` is also a variable
- `new_sp_m014` ... are likely to be values, not variables

```{r}
who1 <- who %>%
  gather(new_sp_m014:newrel_f65, key="key", value="cases", na.rm = TRUE)
who1
```

We can get some hint of the structure of the values in the new `key` column by counting them

```{r}
who1 %>% count(key)
```

We can have the data dictionary handy:

1. The first three letters of each column denote whether the column contains new or old cases of TB
2. The next two letter describe the type of TB:
* `rel` stands for cases of relapse
* `ep` stands for cases of extrapulmonary TB
* `sn` stands for cases of pulmonary TB that could not be diagnosed by a pulmonary smear
* `sp` stands for cases of pulmonary TB that could be diagnosed by a pulmonary smear
3. The sixth letter gives the sex of TB patients
4. The remaining numbers gives the age group. The dataset groups cases into 7 age groups:
* 014 = 0 - 14 years old
* 1524 = 15 - 24 years old
* 2534 = 25 - 34 years old
* 3544 = 35 - 44 years old
* 4554 = 45 - 54 years old
* 5564 = 55 - 64 years old
* 65 = 65 or older

The names of variables are slightly inconsistent so we need to fix them (`new_rel` instead of `newrel`)

```{r}
who2 <- who1 %>%
  mutate(key = stringr::str_replace(key, "newrel", "new_rel"))
who2
```

We can now separate the values in each code

```{r}
who3 <- who2 %>%
  separate(key, c("new", "type", "sexage"), sep = "_")
who3
```

Then we might drop the `new` column because it's constant in this dataset. Let's drop `iso2` and `iso3` too since they are redundant.

```{r}
who4 <- who3 %>%
  select(-new, -iso2, -iso3)
```

Next we'll separate `sexage` into `sex` and `age by splitting after the first character

```{r}
who5 <- who4 %>%
  separate(sexage, c("sex", "age"), sep=1)
who5
```

We can build the whole tidy process in a long pipeline

```{r}
library(tidyverse)
tidywho <- who %>%
  gather(new_sp_m014:newrel_f65, key="key", value="cases", na.rm = TRUE) %>%
  mutate(key = stringr::str_replace(key, "newrel", "new_rel")) %>%
  separate(key, c("new", "type", "sexage"), sep = "_") %>%
  select(-new, -iso2, -iso3) %>%
  separate(sexage, c("sex", "age"), sep=1)
```

```{r}
cyswho <- tidywho %>% group_by(country, year, sex) %>% summarise(c = sum(cases))
ggplot(data = cyswho) +
  geom_bar(mapping = aes(x = year, y = c, fill = sex), stat = "identity") + 
  facet_wrap(~ country, nrow = 10)
```

```{r}
top_country <- tidywho %>% 
  group_by(country) %>% 
  summarise(c = sum(cases)) %>%
  arrange(desc(c))
ggplot(data = cyswho %>% filter(country %in% top_country$country[1:10])) +
  geom_bar(mapping = aes(x = year, y = c, fill = sex), stat = "identity") + 
  facet_wrap(~ country, nrow = 5)
```

